Reducing Linearizability of Priority Queues and More Data Structures to State Reachability

### 2017-04-22 checking-executions.tex
example

We explain how the procedure works on the following execution:
\begin{align}
\hspace{-5mm}\textit{put}(c,p_2) \cdot \textit{put}(a,p_1) \cdot \textit{rm}(a) \cdot \textit{rm}(c) \cdot \textit{rm}(\textit{empty}) \cdot \textit{put}(d,p_2) \cdot \textit{put}(e,p_3) \cdot \textit{rm}(e) \cdot \textit{put}(b,p_1)\label{eq:ex_rec1}
\end{align}
where $p_1$, $p_2$, $p_3$ are priorities such that $p_1 \prec p_2$ and $p_1 \prec p_3$, and $p_2$ and $p_3$ are incomparable. Since the $\textit{rm}(\textit{empty})$ operations\footnote{An $m(a)$-operation in an execution $e$ is an operation identifier $o$ s.t. $e$ contains the actions $\textit{call}_o(m,a)$, $\textit{ret}_o(m,a)$.} are read-only (they don't affect the state of the queue), they are selected first. Ensuring that an operation $o=\textit{rm}(\textit{empty})$ is correct boils down to checking that every $\textit{put}(x,p)$ operation before $o$ is matched to a $\textit{rm}(x)$ operation which also occurs before $o$. This is true in this case for $x\in \{a,c\}$. Therefore, the correctness of (\ref{eq:ex_rec1}) reduces to the correctness of
\begin{align*}
\textit{put}(c,p_2) \cdot \textit{put}(a,p_1) \cdot \textit{rm}(a) \cdot \textit{rm}(c) \cdot \textit{put}(d,p_2) \cdot \textit{put}(e,p_3) \cdot \textit{rm}(e) \cdot \textit{put}(b,p_1)
\end{align*}
When there are no more $\textit{rm}(\textit{empty})$ operations, the procedure selects a $\textit{put}$ operation adding a value with maximal priority which is not removed, and then a pair of $\textit{put}$ and $\textit{rm}$ operations adding and removing the same maximal priority value. For instance, since $p_2$ is a maximal priority, it may select the operation $\textit{put}(d,p_2)$. {This operation is correct if $d$ is the last value with priority $p_2$, and the correctness of (\ref{eq:ex_rec1}) reduces to the correctness of
\begin{align*}
\textit{put}(c,p_2) \cdot \textit{put}(a,p_1) \cdot \textit{rm}(a) \cdot \textit{rm}(c) \cdot \textit{put}(e,p_3) \cdot \textit{rm}(e) \cdot \textit{put}(b,p_1)
\end{align*}
Since there is no other value of priority $p_2$ which is not removed, the procedure selects a pair of $\textit{put}$/$\textit{rm}$ operations with an argument of priority $p_2$,} for instance, $\textit{put}(c,p_2)$ and $\textit{rm}(c)$. The value returned by $\textit{rm}$ is correct if all the values of priority smaller than $p_2$ added before $\textit{rm}(c)$ are also removed before $\textit{rm}(c)$. In this case, $a$ is the only value of priority smaller than $p_2$ and it satisfies this property. Applying a similar reasoning for all the remaining values, it can be proved that this execution is correct.

@@

The following lemma states the correctness of $\textit{Check-PQ-Seq}$ (the proof can be found in Appendix~\ref{sec:appendix definition of seqPQ and proof of Lemma EQP rules and semantics}).

@@

Let us briefly explain the idea of proving this property when $\mathsf{MatchedMaxPriority}$ holds with an example. Assume $e \sqsubseteq l=u \cdot \textit{put}(x,p) \cdot v \cdot \textit{rm}(x) \cdot w$, $\mathsf{MatchedMaxPriority}(l,x)$ holds, and $e \setminus x \sqsubseteq l' \in \textit{PQ}$, as shown in \figurename~\ref{fig:concurrent execution for EPQ1} (a), we explicitly construct a sequence $l''= l''_1 \cdot \textit{put}(x,p) \cdot l''_2 \cdot \textit{rm}(x) \cdot l''_3$ and prove that $e \sqsubseteq l'' \in \textit{PQ}$. In this example, $p_1 \prec p$, $p_1 \prec p_2$, $u=\epsilon$, $w$ contains $\textit{rm}(z_2), \textit{put}(x_3,p_2), \textit{rm}(z_1)$ (emphasize by $\textit{rm}(z_2)-w$), and the remanning operations are in $v$. We also explicitly draw the linearization points according to $l'$. We construct $l''$ as follows, and this method does not rely on the locations of linearization points of $l'$ and operations of $x$.

\begin{itemize}
\setlength{\itemsep}{0.5pt}
\item[-] Let $p$-comparable operations (resp., $p$-incomparable operations) be the set of operations with items whose priority is comparable with $p$ (resp., incomparable with $p$). It seems that we can construct $l''_1$, $l''_2$ and $l''_3$ as the projection of $l'$ into operations of $u$, $v$ and $w$, respectively. However, this is incorrect, since $\mathsf{MatchedMaxPriority}(l,x)$ has no restriction to $p$-incomparable operations operations in $u \cdot v$, and thus, there is no guarantee that the projection of $l'$ into $p$-incomparable operations operations in $u \cdot v$ being correct. In this example, such projection is $\textit{put}(z_1,p_3) \cdot \textit{rm}(z_2)$ and is incorrect.

\item[-] Let us construct sets $U'$, $V'$ and $W'$, such then make $l''_1$, $l''_2$ and $l''_3$ the projections of $l'$ into $U'$, $V'$ and $W'$, respectively. This contains two steps:

\item[-] The first step is to define $W'$. The $p$-comparable operations in $W'$ is same as that in $w$. To obtain $p$-incomparable operations in $W'$, we try to find a $p$-incomparable operation $o$ which either happens before some $p$-comparable operations in $w$, or with linearization points after $\textit{rm}(x)$. Then, we put $o$ and all $p$-incomparable operations after $o$ in $l'$ into $W'$ (emphasized by boxes in example). In this example, $o$ is $\textit{rm}(z_1)$, and $p$-incomparable operations after $o$ in $l'$ is $\textit{rm}(z_2)$, which are emphasized by adding boxes. In this process, whether a $p$-incomparable operation is in $W'$ or not only relies whether it is before or after $o$ in $l'$.

\item[-] The second step is to define $U'$ and $V'$. $U'$ contains the following two kinds of operations: (1) Operations whose linearization points are before $\textit{ret}(\textit{put},x)$, and (2) other $\textit{put}$ operations with priority $p$. $V'$ contains the remanning operations. In this example, $U'$ contains $\textit{put}(z_1,p_3)$ and $\textit{put}(x_2,p_2)$.

\item[-] In this example, $l''_1 = \textit{put}(z_1,p_2) \cdot \textit{put}(x_2,p)$, $l''_2 = \textit{put}(z_2,p_2) \cdot \textit{rm}(x_2) \cdot \textit{put}(y_1,p_1) \cdot \textit{rm}(y_1)$, and $l''_3 = \textit{put}(x_3,p) \cdot \textit{rm}(z_1) \cdot \textit{rm}(z_2)$. In \figurename~\ref{fig:concurrent execution for EPQ1} (b), we add linarization points according to $l''$, and we can see that $l''$ holds as required.
\end{itemize}

@@

%This completes the proof of this lemma.
%
%
%\qed
%\end {proof}

\begin{figure}[t]
  \centering
  \includegraphics[width=1 \textwidth]{figures/PIC-HIS-EPQ1-TwoHis.pdf}
%\vspace{-10pt}
  \caption{The process of obtaining linearization of $e$}
  \label{fig:concurrent execution for EPQ1}
\end{figure}

>>>>

We explain how the procedure works on the following execution:
\begin{align}
\hspace{-5mm}\textit{put}(c,p_2) \cdot \textit{put}(a,p_1) \cdot \textit{rm}(a) \cdot \textit{rm}(c) \cdot \textit{rm}(\textit{empty}) \cdot \textit{put}(d,p_2) \cdot \textit{put}(e,p_3) \cdot \textit{rm}(e) \cdot \textit{put}(b,p_1)\label{eq:ex_rec1}
\end{align}
where $p_1$, $p_2$, $p_3$ are priorities such that $p_1 \prec p_2$ and $p_1 \prec p_3$, and $p_2$ and $p_3$ are incomparable. Since the $\textit{rm}(\textit{empty})$ operations\footnote{An $m(a)$-operation in an execution $e$ is an operation identifier $o$ s.t. $e$ contains the actions $\textit{call}_o(m,a)$, $\textit{ret}_o(m,a)$.} are read-only (they don't affect the state of the queue), they are selected first. Ensuring that an operation $o=\textit{rm}(\textit{empty})$ is correct boils down to checking that every $\textit{put}(x,p)$ operation before $o$ is matched to a $\textit{rm}(x)$ operation which also occurs before $o$. This is true in this case for $x\in \{a,c\}$. Therefore, the correctness of (\ref{eq:ex_rec1}) reduces to the correctness of
\begin{align*}
\textit{put}(c,p_2) \cdot \textit{put}(a,p_1) \cdot \textit{rm}(a) \cdot \textit{rm}(c) \cdot \textit{put}(d,p_2) \cdot \textit{put}(e,p_3) \cdot \textit{rm}(e) \cdot \textit{put}(b,p_1)
\end{align*}
When there are no more $\textit{rm}(\textit{empty})$ operations, the procedure selects a $\textit{put}$ operation adding a value with maximal priority which is not removed, and then a pair of $\textit{put}$ and $\textit{rm}$ operations adding and removing the same maximal priority value. For instance, since $p_2$ is a maximal priority, it selects the operation $\textit{put}(d,p_2)$. {This operation is correct if $d$ is the last value with priority $p_2$, and the correctness of (\ref{eq:ex_rec1}) reduces to the correctness of
\begin{align*}
\textit{put}(c,p_2) \cdot \textit{put}(a,p_1) \cdot \textit{rm}(a) \cdot \textit{rm}(c) \cdot \textit{put}(e,p_3) \cdot \textit{rm}(e) \cdot \textit{put}(b,p_1)
\end{align*}
Since there is no other value of maximal priority which is not removed, the procedure selects a pair of $\textit{put}$/$\textit{rm}$ operations with an argument of maximal priority $p_2$,} for instance, $\textit{put}(c,p_2)$ and $\textit{rm}(c)$. The value returned by $\textit{rm}$ is correct if all the values of priority smaller than $p_2$ added before $\textit{rm}(c)$ are also removed before $\textit{rm}(c)$. In this case, $a$ is the only value of priority smaller than $p_2$ and it satisfies this property. Applying a similar reasoning for all the remaining values, it can be proved that this execution is correct.

@@

The following lemma states the correctness of $\textit{Check-PQ-Seq}$ (see Appendix~\ref{sec:appendix definition of seqPQ and proof of Lemma EQP rules and semantics} for the proof).

@@

We explain the proof of this property on the execution $e$ in \figurename~\ref{fig:concurrent execution for EPQ1}(a) where $p_1 \prec p$, $p_1 \prec p_2$, and the predicate $\mathsf{Has\text{-}MatchedMaxPriority}(e)$ holds. Assume that there exist two sequential executions $l$ and $l'$ such that $e \sqsubseteq l=u \cdot \textit{put}(x,p) \cdot v \cdot \textit{rm}(x) \cdot w$, $\mathsf{MatchedMaxPriority\text{-}Seq}(l,x)$ holds, and $e \setminus x \sqsubseteq l' \in \seqPQ$. Let $u=\epsilon$, $w$ be any sequence containing the set of operations $\textit{rm}(z_2)$, $\textit{put}(x_3,p_2)$, and $\textit{rm}(z_1)$ (we distinguish them by adding the suffix $-w$ to their name, e.g., $\textit{rm}(z_2)-w$), and $v$ be any sequence containing the remaining operations. In general, the linearization $l'$ can be defined by choosing for each operation, a point in time between its call and return actions, called \emph{linearization point}. The order between the linearization points defines the sequence $l'$. \figurename~\ref{fig:concurrent execution for EPQ1}(a) draws linearization points for the operations in $e \setminus x$ defining the linearization $l'$~\footnote{In general, there may exist multiple ways of choosing linearization points to define the same linearization. Our construction is agnostic to this choice.}.
%(the order of these linearization points defines the sequence $l'$). 
We show how to construct a sequence $l''= l''_1 \cdot \textit{put}(x,p) \cdot l''_2 \cdot \textit{rm}(x) \cdot l''_3\in\seqPQ$ such that $e \sqsubseteq l''$. 
%In this example, . We also explicitly draw the linearization points according to $l'$. We construct $l''$ as follows, and this method does not rely on the locations of linearization points of $l'$ and operations of $x$.
\begin{itemize}
\setlength{\itemsep}{0.5pt}
\item[-] An operation is called $p$-comparable (resp., $p$-incomparable) when it receives as argument a value of priority comparable to $p$ (resp., incomparable to $p$). We could try to define $l''_1$, $l''_2$ and $l''_3$ as the projection of $l'$ to the operations in $u$, $v$ and $w$, respectively. However, this is incorrect, since $\mathsf{MatchedMaxPriority\text{-}Seq}(l,x)$ imposes no restriction on  $p$-incomparable operations in $u \cdot v$, and thus, there is no guarantee that the projection of $l'$ to $p$-incomparable operations in $u \cdot v$ is correct. In this example, this projection is $\textit{put}(z_1,p_3) \cdot \textit{rm}(z_2)$ which is incorrect.

\item[-] We define the sets of operations $U'$, $V'$ and $W'$ such that $l''_1$, $l''_2$ and $l''_3$ are the projections of $l'$ to $U'$, $V'$, and $W'$, respectively. This is done in two steps:

\item[-] The first step is to define $W'$. The $p$-comparable operations in $W'$ are the same as in $w$. To identify the $p$-incomparable operations in $W'$, we search for a $p$-incomparable operation $o$ which either happens before some $p$-comparable operation in $w$, or whose linearization point occurs after $\textit{rm}(x)$. We add to $W'$ the operation $o$ and all the $p$-incomparable operations occurring after $o$ in $l'$. In this example, $o$ is $\textit{rm}(z_1)$ and the only $p$-incomparable operation occurring after $o$ in $l'$ is $\textit{rm}(z_2)$ (they are surrounded by boxes in the figure). In this process, whether a $p$-incomparable operation is in $W'$ or not only relies on whether it is before or after $o$ in $l'$.

\item[-] The second step is to define $U'$ and $V'$. $U'$ contains two kinds of operations: (1) operations whose linearization points are before $\textit{ret}(\textit{put},x)$, and (2) other $\textit{put}$ operations with priority $p$. $V'$ contains the remaining operations. In this example, $U'$ contains $\textit{put}(z_1,p_3)$ and $\textit{put}(x_2,p_2)$.

\item[-] In conclusion, we have that $l''_1 = \textit{put}(z_1,p_2) \cdot \textit{put}(x_2,p)$, $l''_2 = \textit{put}(z_2,p_2) \cdot \textit{rm}(x_2) \cdot \textit{put}(y_1,p_1) \cdot \textit{rm}(y_1)$, and $l''_3 = \textit{put}(x_3,p) \cdot \textit{rm}(z_1) \cdot \textit{rm}(z_2)$. \figurename~\ref{fig:concurrent execution for EPQ1}(b) draws linearization points for each operation in $e$ defining the linearization $l''$.
\end{itemize}

@@

%This completes the proof of this lemma.
%
%
%\qed
%\end {proof}
\begin{figure}[t]
  \centering
  \includegraphics[width=1 \textwidth]{figures/PIC-HIS-EPQ1-TwoHis.pdf}
%\vspace{-10pt}
  \caption{The process of obtaining linearization of $e$}
  \label{fig:concurrent execution for EPQ1}
\end{figure}
